An Evolutionary Algorithm for the Multiobjective Risk-Equity Constrained Routing Problem

### 2011-11-16 DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
added a bit more content and ideas

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% our packages and definitions: %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{url} 
\usepackage{color}
% \usepackage{dsfont}

@@

Let the transportation network be represented as a directed graph $G = (N, A)$, with $N$ being the set of nodes and $A$ the set of arcs. Let $C$ be the set of commodities, given as a set of point-to-point demands to transport a certain amount of hazmats. For any commodity $c\in C$, let $s^c$ and $t^c$ be the source node and the destination node respectively, and let $V^c$ be the number of available trucks for the transportation of commodity $c$. Each commodity is associated with a unique type of hazmat. We assume that the risk is computed on each arc of the network and is proportional to the number of trucks traversing such an arc. Although such a proportionality is not always given in pracice, our assumption is a good approximation if the trucks are of the same type and/or the load is very similar as it is the case for standardized castor containers for radioactive material\footnote{Note that changing the proposed model towards different types of trucks with different types of loads and therefore different resulting risks is straightforward and is expected to cause no additional difficulties for the proposed evolutionary algorithm.}. With each arc $(i,j) \in A$, we also associate a cost $c_{ij}^c$, induced by the travel of a truck of commodity $c$ on this arc.

@@

\subsection{An optimization model} \label{SS_AOM}
We introduce a flow variable $f_{ij}^c$ defining the portion of commodity $c$ being transported on arc $(i,j)$. These variables are subject to flow conservation constraints:
\[
	\sum_{j \in \delta^+(i)} f_{ij}^c -  \sum_{j \in \delta^-(i)} f_{ji}^c = b_i^c, \forall i \in N,  c \in  C
\]
where $\delta^-(i)$ and $\delta^+(i)$ are the forward and backward star of $i$, i.e., $\delta^-(i) = \{j \in N: (j,i)\in A\}$ and $\delta^+(i) = \{j \in N: (i,j)\in A\}$, and $b_i^c=\left\{
\begin{array}{rl}
                     1 & \textrm{ if } i=s^c\\
                    -1 & \textrm{ if } i=t^c\\
                     0 & \textrm{ otherwise.}\\
\end{array}\right.$

The first objective is a cost function given as follows:

$$\min \sum_{c \in C} \sum_{(i,j) \in A} c_{ij} f_{ij}^c$$

The second objective is the minimization of the total risk

$$
\min \sum_{c \in C} \sum_{(i,j) \in A} r_{ij}^c f_{ij}^c
$$

We define the risk $w_{ij}$ imposed on the population around the arc $(i,j)$ as a linear combination of the flow variables:

$$w_{ij} = \sum_{c \in C} r_{ij}^c f_{ij}^c$$

and add a new variable $z = \max_{(i,j) \in A} w_{ij}$, which therefore is subject to the constraints

$$
z \geq \sum_{c \in C} r_{ij}^c f_{ij}^c, \forall (i,j) \in A
$$

% We define the risk $w_p$ imposed on the population around the path $p$ as a linear combination of the flow variables: 
% 
% $$w_p = \sum_{c \in C} \sum_{(i,j) \in p} r_{ij}^c f_{ij}^c$$
% 
% and add a new variable $y = \max_{p \in \mathcal{P}} w_p$, which therefore is subject to the constraints
% 
% $$
% y \geq \sum_{c \in C} \sum_{(i,j) \in p} r_{ij}^c f_{ij}^c, \forall p \in \mathcal{P}
% $$
% 
% where $\mathcal{P}$ is the set of paths.

% 
% 
% 
% The third and fourth objectives will then be:
% $$\min \max_{(i,j)\in A} r_{ij} f_{ij}^c$$, and 
% $$\min \max_{p \in \mathcal{P}} \sum_{(i,j) \in p} r_{ij} f_{ij} $$ 
% 
% where

% We introduce the integer variable $y_{ij}^c$ for the number of trucks that use arc $(i,j)$ for transporting commodity $c$. We assume a fixed number of trucks and that all trucks have the same load.

The proposed model is defined as follows:
\begin{eqnarray}
   \min  &  \sum_{c \in C} \sum_{(i,j) \in A} c_{ij} f_{ij}^c & \label{eq:obj1}\\
   \min  &  \sum_{c \in C} \sum_{(i,j) \in A} r_{ij}^c f_{ij}^c & \label{eq:obj2}\\
   \min  & z & \label{eq:obj3}\\
   %\min  & y & \label{eq:obj4}\\
    s.t. & \sum_{j \in \delta^+(i)} f_{ij}^c -  \sum_{j \in \delta^-(i)} f_{ji}^c = d_i^c\quad         & \forall i \in N,  c \in  C \label{eq:flow} \\
    & z \geq \sum_{c \in C} r_{ij}^c f_{ij}^c & \forall (i,j) \in A \\
%     & y \geq \sum_{c \in C} \sum_{(i,j) \in p} r_{ij}^c f_{ij}^c & \forall p \in \mathcal{P}
%          & y_{ij}^c \in \{0, 1, 2, \ldots\}                                                            & \forall (i,j)\in A, c\in C\label{eq:integerY}.
\end{eqnarray}

% The first objective in (\ref{eq:obj1}) is a cost function and is to be minimized. The second objective, given by (\ref{eq:obj2}), minimizes the total risk on all regions and objective (\ref{eq:obj3}) minimizes the maximum risk imposed on all regions. Constraints (\ref{eq:flow}) are conservation constraints, where $\delta^-(i) = \{j \in N: (j,i)\in A\}$ and $\delta^+(i) = \{j \in N: (i,j)\in A\}$ are the direct successors and predecessors of node $i$, and $q_i^c=V^c$ if  $i=s^c$, $q_i^c=-V^c$ if  $i=t^c$ and $q_i^c=0$ otherwise.

% \begin{Rq}
% The number of constaints $nbC = |N| + |A| + |\mathcal{P}|$ is exponontial, so the classical optimization softwares such as CPLEX can not be used.
% \end{Rq}

\section{Evolutionary multiobjective optimization}
\label{S_MEO}
Evolutionary algorithms (EAs) and Evolutionary Multiobjective Optimization (EMO) algorithms in particular are general-purpose randomized search heuristics and as such well suited for problems where the objective function(s) can be highly non-linear, noisy, or even given only implicitly, e.g., by expensive simulations \citep{deb2001a,cvl2007a}. Since the third objective in the above problem formulation is nonlinear, we propose to use an EMO algorithm for the multiobjective risk-equity constrained routing problem here. Our EMO algorithm follows the standard iterative/generational cycle of an EA of mating selection, variation, objective function evaluation, and environmental selection and is build upon the state-of-the-art selection scheme in HypE \citep{bz2011a} as implemented in the PISA platform \citep{bltz2003a}. The variation operators as well as the representation of the solutions, however, have to be adapted to the problem at hand in the following way in order to fulfill the problem's constraints at all times\footnote{The source code of the corresponding PISA hazmat variator is going to be available on the official PISA web page \url{http://www.tik.ee.ethz.ch/pisa/} and until then can be downloaded on the preliminary web page \url{http://researchers.lille.inria.fr/~brockhof/downloads/hazmatvariator20111110.zip}}.

@@

\subsection{Implementation in PISA}
The proposed algorithm has been implemented within the multiobjective optimization framework PISA \citep{bltz2003a} and is going to be made available via the official PISA web page\footnote{Until the source code is available via the PISA web page \url{http://www.tik.ee.ethz.ch/pisa/}, we make it temporarily available at \url{http://researchers.lille.inria.fr/~brockhof/downloads/hazmatvariator20111110.zip} as supplementary material for the review process.}.

\section{Experimental analysis}

We concentrated our analysis on a real-world case study \cite{bianco09}. We considered the road network of the Lazio region (located in the middle of Italy), and, in particular, its main transport roads. The network is composed of 311 nodes and 441 links. The risk $r_{ij}^c, \forall (i,j) \in A, c \in C$ is evaluated as the societal risk computed as the number of people living inside the exposure zone around link $(i,j)$ (whose size depends on the hazmat type $c$ times the accident probability involving the vehicle. We considered from 2 to 10 origin-destination pairs on the network, each one associated with a number of commodities ranging from 1 to 3, for an overall number of shipments (commodities) between 2 and 30 (we recall that in our model each commodity is associated with one origin-destination pair). With each one of the latter scenarios we associated 10 instances; each instance has been generated by assigning uniformly at random a demand from 100 to 1000 tons to each shipment.

@@

\section{Conclusions}
\label{S_FW}
The transportation of hazmats is an important optimization problem in the field of sustainable development and in particular the equitable distribution of risks is of high interest. Within this study, we formalize this transportation problem as the minimization of three objectives and propose to use an evolutionary algorithm to cope with the non-linear equity risk objective.

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% our packages and definitions: %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{url} 
\usepackage{color}
% \usepackage{dsfont}
\usepackage{booktabs}

@@

Let the transportation network be represented as a directed graph $G = (N, A)$, with $N$ being the set of nodes and $A$ the set of arcs. Let $C$ be the set of commodities, given as a set of point-to-point demands to transport a certain amount of hazmats. For any commodity $c\in C$, let $s^c$ and $t^c$ be the source node and the destination node respectively, and let $V^c$ be the number of available trucks for the transportation of commodity $c$. Each commodity is associated with a unique type of hazmat. We assume that the risk is computed on each arc of the network and is proportional to the number of trucks traversing such an arc. Although such a proportionality is not always given in pracice, our assumption is a good approximation if the trucks are of the same type and/or the load is very similar as it is the case for standardized castor containers for radioactive material\footnote{Note that changing the proposed model towards different types of trucks with different types of loads and therefore different resulting risks is straightforward and is expected to cause no additional difficulties for the proposed evolutionary algorithm.}. With each arc $(i,j) \in A$, we also associate a cost $c_{ij}$, induced by the travel of a truck on this arc.

@@

\subsection{The optimization model} \label{SS_AOM}
We introduce the integer variable $y_{ij}^c$ for the number of trucks that use arc $(i,j)\in A$ for transporting commodity $c$. As mentioned above, we assume a fixed number of trucks for each commodity and that all trucks associated to a certain commodity have the same load. The proposed model is defined as follows:
\begin{eqnarray}
   \min  & \sum_{c \in C}\limits \sum_{(i,j) \in A}\limits c_{ij} y_{ij}^c & \label{eq:obj1}\\
   \min  & \sum_{c \in C}\limits \sum_{(i,j) \in A}\limits r_{ij}^{c} y_{ij}^c & \label{eq:obj2}\\
   \min  & \max_{(i,j)\in A}\limits \big\{ \sum_{c \in C}\limits r_{ij}^{c} y_{ij}^c \big\} & \label{eq:obj3}\\
    s.t. & \sum_{j \in \delta^+(i)} y_{ij}^c -  \sum_{j \in \delta^-(i)} y_{ji}^c = q_i^c\quad         & \forall i \in N,  c \in  C \label{eq:flow} \\
         & y_{ij}^c \in \{0, 1, 2, \ldots\}                                                            & \forall (i,j)\in A, c\in C\label{eq:integerY}.
\end{eqnarray}

The first objective in (\ref{eq:obj1}) is a cost function and is to be minimized. The second objective, given by (\ref{eq:obj2}), minimizes the total risk over all arcs and objective (\ref{eq:obj3}) minimizes the maximum risk imposed on all arcs. The constraints in (\ref{eq:flow}) are conservation constraints, where $\delta^-(i) = \{j \in N: (j,i)\in A\}$ and $\delta^+(i) = \{j \in N: (i,j)\in A\}$ are the direct successors and predecessors of node $i$, and $q_i^c=V^c$ if  $i=s^c$, $q_i^c=-V^c$ if  $i=t^c$ and $q_i^c=0$ otherwise. Note that while the first two objectives are linear in the variables, the maximum in the third objective makes it highly non-linear---the main reason why we propose to use an evolutionary algorithm to deal with the above hazmat routing problem.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Evolutionary multiobjective optimization} \label{S_MEO}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Evolutionary algorithms (EAs) and Evolutionary Multiobjective Optimization (EMO) algorithms in particular are general-purpose randomized search heuristics and as such well suited for problems where the objective function(s) can be highly non-linear, noisy, or even given only implicitly, e.g., by expensive simulations \citep{deb2001a,cvl2007a}. Since the third objective in the above problem formulation is nonlinear, we propose to use an EMO algorithm for the multiobjective risk-equity constrained routing problem here. Our EMO algorithm follows the standard iterative/generational cycle of an EA of mating selection, variation, objective function evaluation, and environmental selection and is build upon the state-of-the-art selection scheme in HypE \citep{bz2011a} as implemented in the PISA platform \citep{bltz2003a}. The variation operators as well as the representation of the solutions, however, have to be adapted to the problem at hand in the following way in order to fulfill the problem's constraints at all times\footnote{The source code of the corresponding PISA hazmat variator is going to be available on the official PISA web page \url{http://www.tik.ee.ethz.ch/pisa/} and until then can be downloaded on the preliminary web page \url{http://researchers.lille.inria.fr/~brockhof/downloads/hazmatvariator20111110.zip}}. Moreover, two different initialization schemes are proposed to deal with the specific variable length representation.

@@

\subsection{Implementation in PISA}
The proposed algorithm has been implemented within the multiobjective optimization framework PISA \citep{bltz2003a} and is going to be made available via the official PISA web page\footnote{Until the source code is available via the PISA web page \url{http://www.tik.ee.ethz.ch/pisa/}, we made it temporarily available at \url{http://researchers.lille.inria.fr/~brockhof/downloads/hazmatvariator20111110.zip} as supplementary material for the review process.}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Experimental analysis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The following experiments have two goals:
- show that the algorithm works in principle on practical examples in manageable time
- check which of the two initialization schemes is 
In addition, a few more experiments show that the standard parameters of the algorithm behaves as with other test problems: population size, environmental selection (change from HypE to NSGA-II)

As a test case, we chose to use a real-world case study based on the work of \citet{bianco09}. The network used here consists of 311 nodes and 441 arcs and represents the main transport roads in Italy's Lazio region which amongst other cities includes the capital Rome.

The risks $r_{ij}^c$ for each arc $(i,j) \in A$ and commodity $c \in C$ are evaluated as the societal risk computed as the number of people living inside the exposure zone around link $(i,j)$ (whose size depends on the hazmat type $c$ times the accident probability involving the vehicle), for details see  \citep{bianco09}.

%We concentrated our analysis on a real-world case study \citep{bianco09}. We considered the road network of the Lazio region (located in the middle of Italy), and, in particular, its main transport roads. The network is composed of 311 nodes and 441 links. The risks $r_{ij}^c$ for each arc $(i,j) \in A$ and commodity $c \in C$ are evaluated as the societal risk computed as the number of people living inside the exposure zone around link $(i,j)$ (whose size depends on the hazmat type $c$ times the accident probability involving the vehicle), for details see  \citep{bianco09}. We considered from 2 to 10 origin-destination pairs on the network, each one associated with a number of commodities ranging from 1 to 3, for an overall number of shipments (commodities) between 2 and 30 (we recall that in our model each commodity is associated with one origin-destination pair). With each one of the latter scenarios we associated 10 instances; each instance has been generated by assigning uniformly at random a demand from 100 to 1000 tons to each shipment.

\subsection{Experimental Settings}
population size: 100, 2000 generations, 10 runs, hypervolume's reference point for HypE as in Table~\ref{tab:referencepoints} chosen according to initial objective function values: such that all initial solutions have a contribution to the hypervolume indicator (ref point the same in all objectives due to off-the-shelf PISA implementation of HypE), hypervolume computation exact

\begin{table}
\centering
\caption{\label{tab:referencepoints} Used reference points within the environmental selection of HypE}
\begin{tabular}{lr}
	\toprule
	instance & reference point\\ \midrule
	\# commodities: 2 & $(10^8, 10^8, 10^8)$\\
	\# commodities: 3 & $(10^8, 10^8, 10^8)$\\
	\# commodities: 4 & $(10^9, 10^9, 10^9)$\\
	\bottomrule
\end{tabular}
\end{table}

@@

\subsection{The Influence of Standard Parameters}
\begin{itemize}
	\item show exemplarily the influence of some of the standard parameters of the EMO algorithm which 
\end{itemize}

\subsection{The Feasibility of the EMO Approach}
\begin{itemize}
	\item of course no guarantee that the solutions are Pareto-optimal
	\item however, in reasonable time (e.g. ??hrs for 10 runs with 1000 generations each) results with several non-dominated solutions the decision maker can interpret and with which she can get an idea about the trade-offs between the objectives
	\item if DM wants good results fast, cost-efficient initialization is favourable, otherwise, random initialization gives better spread of the solutions
	\item if needed, approach can be combined easily with preference articulation techniques such as the weighted hypervolume indicator which allows to steer the search towards user-defined regions of the objective space (show 1 plot with W-HypE).
	\item approach is very general and can easily deal with slightly modified problems: e.g. different truck loads (only the objective function code needs to be changed)
	\item it needs to be mentioned that for the tested instances with 311 nodes and 441 arcs, the variation operator of the evolutionary algorithm was able to construct the shortest paths from source to destination in some of the runs for some of the truck paths within the first 100,000 function evaluations. This corresponds to performing random walks of lengths around 20 nodes.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions} \label{S_FW}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The transportation of hazmats is an important optimization problem in the field of sustainable development and in particular the equitable distribution of risks is of high interest. Within this study, we formalize this transportation problem as the minimization of three objectives and propose to use an evolutionary algorithm to cope with the non-linear equity risk objective.

@@

\section{Acknowledgments}
The authors would like to thank Stefano Giordani for providing us with the instances of the Lazio network. Moreover, Dimo Brockhoff would like to acknowledge support from the CNRS-Microsoft chair "`Optimization for Sustainable Development"' during his time at Ecole Polytechnique, Palaiseau, France.
